---
title: "Checklist-based indicators of Invasive Alien Species"
author:
- Damiano Oldoni
- Add your name
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to build indicators of Invasive Alien Species based on checklist data. In particular, the following indicators are taken into account:

1. number of new introductions of alien species per year in Belgium (see GitHub issue [#17](https://github.com/trias-project/pipeline/issues/17))
2. pathways associated with alien species introductions (see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19))
3. cumulative number of alien species (see GitHub issue [#20](https://github.com/trias-project/pipeline/issues/20))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install `trias` package:

```{r install_packages}
devtools::install_github("trias-project/trias@add_functions_checklist_indicators")
```


Load libraries:

```{r load_libraries}
# Tidyverse packages
library(tidyr)
library(dplyr)
library(magrittr)
library(stringr)
library(readr)
# GBIF related packages
library(rgbif)
# project package
library(trias)
#other pckages
library(lazyeval)
```

# Get data

One of the TrIAS goals is to unify seven authoritative checklists. Such unified checklist, public on GBIF (Global Biodiversity Information Facility) will be the start point of this workflow. At the moment, we don't have such checklist. The analysis is based on the data of the following checklists:

1. [Manual of the Alien Plants of Belgium](https://www.gbif.org/dataset/9ff7d317-609b-4c08-bd86-3bc404b77c42)
2. [Inventory of alien macroinvertebrates in Flanders, Belgium](https://www.gbif.org/dataset/289244ee-e1c1-49aa-b2d7-d379391ce265)
3. [Checklist of non-native freshwater fishes in Flanders, Belgium](https://www.gbif.org/dataset/98940a79-2bf1-46e6-afd6-ba2e85a26f9f)
4. [Catalogue of the Rust Fungi of Belgium](https://www.gbif.org/dataset/b043c480-dd36-4f4f-aa82-e188753ff09d)
5. [RINSE - Registry of non-native species in the Two Seas region countries (Great Britain, France, Belgium and the Netherlands)](https://www.gbif.org/dataset/3f5e930b-52a5-461d-87ec-26ecd66f14a3)
6. [RINSE - Pathways and vectors of biological invasions in Northwest Europe](https://www.gbif.org/dataset/1738f272-6b5d-4f43-9a92-453a8c5ea50a)

## Retrieve data in checklist(s) from GBIF

First step is importing checklists data from GBIF. We do it via `trias` function `gbif_get_taxa`

```{r get_taxa}
checklist_info <- c("key", "nubKey", "scientificName", "datasetKey")
taxon <- trias::gbif_get_taxa(
  checklist_keys = c("9ff7d317-609b-4c08-bd86-3bc404b77c42",
                     "289244ee-e1c1-49aa-b2d7-d379391ce265",
                     "98940a79-2bf1-46e6-afd6-ba2e85a26f9f",
                     "b043c480-dd36-4f4f-aa82-e188753ff09d",
                     "3f5e930b-52a5-461d-87ec-26ecd66f14a3",
                     "1738f272-6b5d-4f43-9a92-453a8c5ea50a"),
  origin = "SOURCE", 
  limit = 99999) %>%
  select(one_of(checklist_info))
```

Extract taxonomic information from GBIF taxonomic backbone checklist. Only taxa found in GBIF backbone will be taken into account (see GitHub issue [#21](https://github.com/trias-project/pipeline/issues/21)), so only taxa with a `nubKey`

```{r retrieve_taxonomic_info_backbone}
taxon %<>% filter(is.finite(nubKey))
taxon_keys <- taxon %>% select(key) %>% pull(key)
nub_keys <- taxon %>% distinct(nubKey)
taxon_backbone_info <-  nub_keys %>%
  rowwise() %>%
  do_(interp(~ rgbif::name_usage(key = .$nubKey,
                         return = "data")))
taxon_info <- c("key", "species", "genus", "family", "order",
                "class", "phylum", "kingdom",
                "rank", "speciesKey", "taxonomicStatus", 
                "acceptedKey", "accepted")
taxon_backbone_info %<>%
  select(key,one_of(taxon_info))
combined_taxon_info <- taxon %>% left_join(taxon_backbone_info,
                                           by = c("nubKey" = "key"))
head(combined_taxon_info)
```

### Distribution

Extract distribution information

```{r retrieve_distribution_information}
distribution <- data.frame(key = taxon_keys) %>%
  rowwise() %>%
  do_(interp(~ as.data.frame(rgbif::name_usage(key = .$key,
                                               return = "data",
                                               data = "distribution")))) %>% 
  select(-source)
head(distribution)
```

### Description

Extract description information

```{r retrieve_description_information}
description <- data.frame(key = taxon_keys) %>%
  rowwise() %>%
  do_(interp(~ as.data.frame(rgbif::name_usage(key = .$key,
                                               return = "data",
                                               data = "description"))))
description %<>% select(-c(key, language, source))
description %<>% ungroup()
head(description, n = 10)
```

Contrarily to distribution information, description contains a column  `type` with the following descriptors: 

```{r}
description %>% distinct(type)
descriptors <- description %>% distinct(type) %>% pull()
```
We would like to *tidy* this data.frame, thus having different descriptors as different columns. Unfortunately `dplyr::spread(key, value)` doesn't work due to *duplicates*, i.e. keys with more than one value, for example:

```{r duplicates_issue_spread}
description %>% filter(taxonKey == 141264585 & type == "native range")
```

A more complex strategy has been found:

```{r tidy_description}
change_colname <- function(df, new_colname){
  df %>% 
    rename(!!new_colname := description) %>%
    select(-type)
}

description_tidy <- 
  purrr::map(descriptors, function(x) description %>% filter(type == x)) %>%
  purrr::map2(descriptors, ~ change_colname(.x, .y)) %>%
  purrr::reduce(dplyr::full_join, by = "taxonKey")
head(description_tidy)
```

The chunk code above should be wrapped in a function: Peter Desmet proposed the name `spread_with_duplicates`. It could be put in project package, `trias`, or be part of `tidyr` package eventually. See GitHub issue [#16](https://github.com/trias-project/trias/issues/16).

### Species profile

Extract species profile

```{r retrieve_species_profile_information}
speciesProfiles <- data.frame(key = taxon_keys) %>%
  rowwise() %>%
  do_(interp(~ as.data.frame(rgbif::name_usage(key = .$key,
                                               return = "data",
                                               data = "speciesProfiles"))))
head(speciesProfiles)
```

### Merge data

Merge all retrieved informations

```{r merge_extensions_information}
merged_extensions <- full_join(distribution,
                               description_tidy,
                               speciesProfiles,
                               by = "taxonKey")

merged_info <- left_join(combined_taxon_info, merged_extensions,
                         by = c("key" = "taxonKey"))
```

## Filter invasive alien species

Filter species by:

1. invasion stage (`establishmentMeans` equal to one of `INTRODUCED`,  `NATURALISED`, `INVASIVE`, `ASSISTED COLONISATION`) see GitHub issue [#28](https://github.com/trias-project/pipeline/issues/28#issuecomment-382036510)
2. presence status (`status` not equal to `DOUBTFUL`, `ABSENT`, `EXCLUDED` or `EXTINCT`) see GitHugb issue [#26](https://github.com/trias-project/pipeline/issues/26#issuecomment-382036007)
3. presence in Belgium (`country` equal to `BE`)

```{r filter_by_establishmentMeans}
merged_info %<>%
  filter(establishmentMeans %in% c("INTRODUCED", "NATURALISED", "INVASIVE",
                                   "ASSISTED COLONISATION"),
         !status %in% c("DOUBTFUL", "ABSENT", "EXCLUDED", "EXTINCT"),
         country == "BE")
```

## Tidy data

Some columns should be splitted in order to make the data.frame completely tidy: `temporal` and `pathway`.

### Split column `temporal`

The column `temporal` contains one or two datums

```{r example_temporal, echo = FALSE}
merged_info %>% 
  select(key, temporal) %>% 
  distinct() %>% 
  filter(!is.na(temporal)) %>%
  head()
```

In case there are two datums, we split them in `first_observed` and `last_observed`. If only a datum is present, then it is used for both  `first_observed`and `last_observed` as suggested [here]().

```{r split_temporal}
merged_info %<>% separate(col = temporal,
                          sep = "/",
                          into = c("first_observed", "last_observed"),
                          convert = TRUE,
                          fill = "right",
                          remove = TRUE) %>%
  mutate(last_observed = ifelse(is.na(last_observed),
                          first_observed,
                          last_observed))
```

As example

```{r example_split_temporal, echo = FALSE}
merged_info %>% 
  filter(locality == "Belgium") %>% 
  select(key, first_observed, last_observed) %>%
  distinct() %>%
  filter(!is.na(first_observed) & !is.na(last_observed)) %>%
  head()
```

### Split column `pathway`

The column `pathway` contains a prefix, `cbd_2014_pathway:` and two different pathway levels divided by symbol `_`

```{r pathway_levels_example, echo = FALSE}
merged_info %>% distinct(pathway) %>% filter(!is.na(pathway)) %>% head()
```

We split this column in `pathway_level1` and `pathway_level2`

```{r split_pathway}
merged_info %<>% rowwise() %>%
  mutate(pathway_level1 = 
           str_split_fixed(
             str_split(pathway,
                       pattern = "pathway:")[[1]][2],
             pattern = "_", n = 2)[[1]][1],
         pathway_level2 = 
           str_split_fixed(
             str_split(pathway,pattern = "pathway:")[[1]][2],
             pattern = "_", n = 2)[[2]][1]) %>%
  ungroup() %>%
  select(-pathway) 
```

As example:

```{r example_split_pathway, echo = FALSE}
merged_info %>% 
  distinct(pathway_level1, pathway_level2) %>%
  filter(!is.na(pathway_level1) & !is.na(pathway_level2)) %>%
  head()
```

## Save data output

The information in `merged_info` is ready to be saved in folder `\data\interim`:

```{r save_data_output, eval=FALSE}
write_tsv(x = merged_info, 
          path = "../data/interim/test_data_output_checklist_indicators.tsv")
```

And be used as start point for building indicators.
