---
title: "Get occurrences for list of species"
author:
- Peter Desmet
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to retrieve occurrences for a list of species.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install packages:

```{r install_packages}
devtools::install_github("trias-project/trias")
devtools::install_github("ropensci/rgbif")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(dplyr)
library(magrittr)
library(readr)

# GBIF package
library(rgbif)
# Project package
library(trias)
```

Set file paths:

```{r set_file_paths}
# All downloads we've triggered are added to this file:
downloads_output_file <- "../data/output/gbif_downloads.tsv"
```

# Define taxa

The list of taxa we want to get occurrences for is maintained in a tsv file. To allow better tracking, the path to the file will be added to the `download_output_file`. Ideally we don't use a local file path, but the URL of a specific commit:

```{r define_taxa}
taxa_input_file <- "https://raw.githubusercontent.com/trias-project/pipeline/occ_indicators_downscaling/data/input/occ_indicator_emerging_species_AOO.tsv"
```

Read file:

```{r read_taxa_file}
taxa <- read_delim(
  file = taxa_input_file, 
  "\t",
  escape_double = FALSE, 
  col_types = cols(entry_into_force = col_date(format = "%Y-%m-%d")), 
  trim_ws = TRUE
)
```

Preview file:

```{r preview taxa}
head(taxa)
```

Retrieve the `taxonKey`s we want to use to download occurrences:

```{r get_taxonkeys}
taxon_keys <- taxa[["backbone_taxonKey"]] # Column with taxonKeys
```

# Define countries

We will limit our occurrences search for specific countries. To allow better tracking, this filter will also be added to `download_output_file`:

```{r define_countries}
countries <- c("BE")
```

# Download GBIF occurrences

**Note**: GBIF credentials are required in the next step.

Trigger download:

```{r trigger_gbif_download}
# Important: query arguments need to be pasts as strings of format "key = value"
gbif_download_key <- occ_download(
  paste0("taxonKey = ", paste(taxon_keys, collapse = ",")), 
  paste0("country = ", paste(countries, collapse = ",")),
  curlopts = list(verbose = TRUE),
  user = rstudioapi::askForPassword("GBIF username"),
  pwd = rstudioapi::askForPassword("GBIF password"),
  email = rstudioapi::askForPassword("Email address for notification")
)
```

If you need only occurrence data with geographic coordinates, you can opt to add additional parameter `hasCoordinate = TRUE` to the query, thus reducing data size.

```r
gbif_download_key <- occ_download(
  paste0("taxonKey = ", paste(taxon_keys, collapse = ",")), 
  paste0("country = ", paste(countries, collapse = ",")),
  paste0("hasCoordinate = TRUE"),
  curlopts = list(verbose = TRUE),
  user = rstudioapi::askForPassword("GBIF username"),
  pwd = rstudioapi::askForPassword("GBIF password"),
  email = rstudioapi::askForPassword("Email address for notification")
)
```

Check status of download:

```{r}
metadata <- occ_download_meta(key = gbif_download_key)
metadata$key
metadata$status
```

Write download to list of downloads and check pending downloads:

```{r}
update_download_list(
  file = downloads_output_file, 
  download_to_add = gbif_download_key, 
  input_checklist = taxa_input_file
)
```

# Import occurrences from an existing GBIF download

To import occurrences from an existing GBIF download, you will need the `gbif_download_key`. This key can be found in the 1st column of [gbif_download.tsv](../data/output/gbif_downloads.tsv), which lists all GBIF downloads triggered for TrIAS. For example:

```{r import_ocurrence_from_gbif_key, evaluate = FALSE}
gbif_key <- "0003400-181003121212138"
file_path <- paste0("../data/output/", gbif_key)
if (!file.exists(file_path))
  dir.create(file_path)
occ <- occ_download_get(key = gbif_key, overwrite = TRUE, path = file_path)
unzip(zipfile = occ, exdir = file_path)
occ_file <- "occurrence.txt"
occurrence <- read_delim(
  paste0(file_path, "/", occ_file), "\t", 
  escape_double = FALSE, 
  trim_ws = TRUE
)
```
