---
title: "Occurrence indicators: pre-processing"
author:
- Damiano Oldoni
- Toon Van Daele
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes the pre-processing of occurrence data of alien species. The outputs of this pipeline will be used as input for modelling.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries:

```{r load_libraries, message = FALSE}
library(tidyverse)      # To do data science
library(tidylog)        # To provide feedback on dplyr functions
library(progress)       # To add progress bars
library(here)           # To find files
library(lubridate)      # To work with dates
library(rgbif)          # To get taxa from publisehd unified checklist
```

# Get data

We get occurrence data as obtained in `trias-project/occ-processing` repository. 

## Alien species 

We load occurrence data related to alien species in Belgium:

```{r read_data_cube}
df <- read_tsv(
  file = "https://raw.githubusercontent.com/trias-project/occ-processing/master/data/processed/cube_belgium.tsv",
  col_types = cols(year = col_double(),
                   eea_cell_code = col_character(),
                   taxonKey = col_double(),
                   n = col_double(),
                   min_coord_uncertainty = col_double()),
  na = ""
)
```

We remove columns we will never use:

```{r remove_coords_uncertainty}
df <- 
  df %>%
  select(-min_coord_uncertainty)
```

## Baseline data

We get data at class level, which will be used for correcting the research bias effort:

```{r get_data_baseline, message = FALSE}
df_bl <- read_tsv(
  file = "https://raw.githubusercontent.com/trias-project/occ-processing/master/data/processed/cube_belgium_baseline.tsv",
  col_types = cols(year = col_double(),
                   eea_cell_code = col_character(),
                   classKey = col_double(),
                   n = col_double(),
                   min_coord_uncertainty = col_double()),
  na = "")
```

We remove columns we will never use:

```{r remove_coords_uncertainty_baseline}
df_bl <- 
  df_bl %>%
  select(-min_coord_uncertainty)
```
## Grid cells intersecting protected areas

We get information about intersection of grid cells with Belgian protected areas by reading file `intersect_EEA_ref_grid_protected_areas.tsv` in  `data/output`:
```{r read_protected_areas}
df_prot_areas <- read_tsv(
  here::here("data", 
       "output",
       "intersect_EEA_ref_grid_protected_areas.tsv"),
  na = ""
)
```

Preview:

```{r preview_prot_areas}
df_prot_areas %>%
  head()
```

We are interested to areas included in Natura2000. We remove columns we are not interested to:

```{r select_cols_prot_areas}
df_prot_areas <-
  df_prot_areas %>%
  select(CELLCODE, 
         natura2000)
```

#  Pre-processing

## Add informative columns

For better interpretation of the data, we retrieve canonical names of the species whose data we are going to analyze. We also retrieve the class of each alien species to link occurrence data with the related baseline data:

```{r get_canonical_names}
taxon_key <- 
  df %>% 
  distinct(taxonKey) %>%
  pull()
pb <- progress_bar$new(total = length(taxon_key))
spec_names <- map_df(
  taxon_key, 
  function(taxonKey){
    pb$tick()
    name_usage(key = taxonKey, return = "data")
  }) %>%
  select(key, canonicalName, kingdomKey, classKey)
```

We also retrieve kingdom and class:

```{r get_kingodm_class}
class_key <- 
  spec_names %>% 
  distinct(classKey) %>%
  filter(!is.na(classKey)) %>%
  pull()
pb <- progress_bar$new(total = length(class_key))
kingdom_class <- map_df(
  class_key,
  function(x){
    pb$tick()
    name_usage(key = x, return = "data")
  }) %>%
  select(classKey, class, kingdomKey, kingdom)
```

and we add them:

```{r add_kingdom_class}
spec_names <-
  spec_names %>%
  left_join(kingdom_class, by = c("kingdomKey", "classKey"))
```

So, these are the covered alien species:

```{r review_species}
spec_names
```

Taxa not assigned to any class:

```{r taxa_no_class}
spec_names %>%
  filter(is.na(classKey))
```

## Extract x-y coordinates from cellcodes

Extract coordinates from column `eea_cell_code`. For example, the coordinates of cellcode `1kmE3895N3116` are `x = 3895` and `y = 3116`:

```{r extract_x_y}
df_xy <- 
  df %>% 
  distinct(eea_cell_code) %>%
  bind_cols(
    tibble(
      x = unlist(str_extract_all(unique(df$eea_cell_code), 
                                 pattern = "(?<=E)\\d+")),
      y = unlist(str_extract_all(unique(df$eea_cell_code), 
                                 pattern = "(?<=N)\\d+"))) %>%
      mutate_all(as.integer)
)
```

Preview:

```{r x_y_preview}
df_xy %>% head()
```

Do the same for baseline data:

```{r extract_x_y_baseline}
df_bl_xy <- 
  df_bl %>% 
  distinct(eea_cell_code) %>%
  bind_cols(
    tibble(
      x = unlist(str_extract_all(unique(df_bl$eea_cell_code), 
                                 pattern = "(?<=E)\\d+")),
      y = unlist(str_extract_all(unique(df_bl$eea_cell_code), 
                                 pattern = "(?<=N)\\d+"))) %>%
      mutate_all(as.integer)
)
```

Preview:

```{r x_y_preview_baseline}
df_bl_xy %>% head()
```

## Apply temporal and spatial restraints

Select observations:

- from 1950, start of state of invasion researches
- to 2017, thus dealing with data publishing delay on GBIF, estimated to be two years behind present date

```{r define_year_range}
first_year <- 1950
last_year <- year(Sys.Date()) - 2
```

Some occurrence data can be published with erroneous coordinates (see [#trias-project/occ-processing#13](https://github.com/trias-project/occ-processing/issues/13)) or can be assigned to a grid cell out of Belgium. We remove these records:

```{r define_x_y_range_values}
be_cellcodes <- df_prot_areas$CELLCODE
```

Apply restraints defined above to both occurrence data:

```{r apply_restraints_time_space_occ_data}
df <- 
  df %>%
  left_join(df_xy, by = "eea_cell_code") %>%
  filter(year > first_year & year <= last_year) %>%
  filter(eea_cell_code %in% be_cellcodes)
```

and baseline data:

```{r apply_restraints_time_space_baseline_data}
df_bl <- 
  df_bl %>%
  left_join(df_bl_xy, by = "eea_cell_code") %>%
  filter(year > first_year & year <= last_year) %>%
  filter(eea_cell_code %in% be_cellcodes)
```

## Add information about presence of protected areas

We add whether the grid cell intersects any of the Natura2000 Belgian protected areas.

```{r add_natura2000_col}
df <-
  df %>%
  left_join(df_prot_areas,
            by = c("eea_cell_code"= "CELLCODE"))
```

Preview:

```{r preview_df_with_natura2000_col}
df %>% head()
```

We do the same for baseline data:

```{r add_natura2000_col_baseline}
df_bl <-
  df_bl %>%
  left_join(df_prot_areas,
            by = c("eea_cell_code"= "CELLCODE"))
```

Preview:

```{r preview_df_bl_with_natura2000_col}
df_bl %>% head()
```

## Calculate number of observations of native species

Calculate the total number of alien species observations per `eea_cell_code`, `year` and `class`:

```{r }
df_alien_species <- 
  df %>%
  left_join(spec_names, by = c("taxonKey" = "key")) %>%
  group_by(classKey, eea_cell_code, year) %>%
  summarize(n_alien = sum(n)) %>%
  ungroup()
df_alien_species
```

Substract the alien species observations from the baseline data to get number of observations of native species:

```{r subtract_alien_species}
df_bl <- 
  df_bl %>%
  left_join(df_alien_species, by = c("classKey", "eea_cell_code", "year")) %>%
  replace_na(list(n_alien = 0)) %>%
  mutate(native_obs = n - n_alien) %>%
  mutate(native_obs = if_else(native_obs < 0, 0, native_obs)) %>%
  select(year, eea_cell_code, classKey, native_obs, natura2000)
```

Preview:

```{r preview_corrected_baseline}
df_bl %>%
  head()
```

## Create time series

For each species, define cells with at least one observation:

```{r distinct_cells_taxa}
df_cc <- 
  df %>%
  group_by(taxonKey) %>%
  distinct(eea_cell_code) %>%
  ungroup()
```

For each species, identify the first year with an observation:

```{r begin_year_per_species}
df_begin_year <- 
  df %>%
  group_by(taxonKey) %>%
  summarize(begin_year = min(year))
```

For each species, combine `begin_year` and unique `eea_cell_code` as found above: 

```{r combine_begin_year_cells_taxa}
df_cc <- 
  df_cc %>%
  left_join(df_begin_year, by = "taxonKey") %>%
  select(taxonKey, begin_year, eea_cell_code)
```

Preview:

```{r preview_begin_year_cells_combo}
df_cc %>% head()
```

For each cell (`eea_cell_code`) and species (`taxonKey`) we can now create a time series from `begin_year` to `last_year`. Notice that at this step we create only time series slots: occurrence data are added at next stage.

```{r make_time_series}
make_time_series <- function(eea_cell_code, taxonKey, begin_year, last_year ) {

    expand_grid(eea_cell_code = eea_cell_code,
                taxonKey = taxonKey,
                year = seq(from = begin_year, to = last_year))

}

df_s <- pmap_dfr(df_cc, .f = make_time_series, last_year = last_year)
```

Fill occurrence data and corrected baseline data to the time series:

```{r add_data}
df_s <- 
  df_s %>%
  # add occurrence data
  left_join(df %>% select(-c(x,y)), 
            by = c("taxonKey", "year", "eea_cell_code")) %>%
  # add classKey
  left_join(spec_names %>% 
              select(key, classKey), 
            by = c("taxonKey" = "key")) %>%
  # add baseline data (at class level)
  left_join(df_bl %>%
              select(year, eea_cell_code, classKey, native_obs),
            by = c("year", "eea_cell_code", "classKey")) %>%
  replace_na(list(native_obs = 0, n = 0))
```

Preview:

```{r}
df_s %>% head(n = 30)
```

For each species, we can now get the time series of number of cells, `ncells`, and number of observations, `obs`.

```{r ts_ncells_nobs}
df_obs <- 
  df_s %>%
  group_by(taxonKey, year) %>%
  summarise(obs = sum(n),
            native_obs = sum(native_obs)) %>%
  ungroup()

# Get number of cells per taxonKey and year
df_ncells <- 
  df_s %>%
  filter(n > 0) %>%
  group_by(taxonKey, year) %>%
  summarise(ncells = n_distinct(eea_cell_code)) %>%
  ungroup()

# Get reference number of cells (native species at class level) per taxonKey and
# year
df_native_ncells <- 
  df_s %>%
  filter(native_obs > 0) %>%
  group_by(taxonKey, year) %>%
  summarise(native_ncells = n_distinct(eea_cell_code)) %>%
  ungroup()

# Merge together
df_ncells_nobs <- 
  df_obs %>%
  left_join(df_ncells,
            by = c("taxonKey", "year")) %>%
  left_join(df_native_ncells,
             by = c("taxonKey", "year")) %>%
  replace_na(list(ncells = 0, native_ncells = 0)) %>%
  # add taxonomic information
  left_join(spec_names,
            by = c("taxonKey" = "key")) %>%
  select(taxonKey, canonicalName, year, everything())
```

Preview:

```{r preview_ncells_nobs}
df_ncells_nobs %>% head(n = 30)
```

# Save data

Save `df_ncells_nobs` as it will be used as start point of modelling pipelines:

```{r save_data_as_tsv}
write_tsv(df_ncells_nobs, 
          path = here::here("data", "interim", "df_n_obs_n_cells.tsv"),
          na = "")
```
