---
title: "Cheklist-based indicators of Invasive Alien Species"
author:
- Damiano Oldoni
- Add your name
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

This document describes how to build indicators of Invasive Alien Species based on checklist data. In particular, the following indicators are taken into account:

1. number of new introductions of alien species per year in Belgium (see GitHub issue [#17](https://github.com/trias-project/pipeline/issues/17))
2. pathways associated with alien species introductions (see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19))
3. cumulative number of alien species (see GitHub issue [#20](https://github.com/trias-project/pipeline/issues/20))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install `trias` package:

```{r install_packages}
devtools::install_github("trias-project/trias")
```


Load libraries:

```{r load_libraries}
# Tidyverse packages
library(dplyr)
library(magrittr)
library(readr)
library(trias)
library(rgbif)
```

# Get data

One of the TrIAS goals is to unify seven authoritative checklists. Such unified cheklist, public on GBIF (Global Biodiversity Information Facility) would be the start point of this workflow. However, we still don't have such checklist. We can still further work based on data of three different checklists:

1. [Manual of the Alien Plants of Belgium](https://www.gbif.org/dataset/9ff7d317-609b-4c08-bd86-3bc404b77c42)
2. [Inventory of alien macroinvertebrates in Flanders, Belgium](https://www.gbif.org/dataset/289244ee-e1c1-49aa-b2d7-d379391ce265)
3. [Checklist of non-native freshwater fishes in Flanders, Belgium](https://www.gbif.org/dataset/98940a79-2bf1-46e6-afd6-ba2e85a26f9f)


