---
title: "Indicator: Number of new introductions of alien species per year"
author:
- Damiano Oldoni
- Yasmine Verzelen
- Stijn Van Hoey
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

This document describes how to build indicators of Invasive Alien Species based 
on checklist data. In particular, this document takes into account:

1. number of new introductions of alien species per year in Belgium (see GitHub 
issue [#17](https://github.com/trias-project/pipeline/issues/17))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install `trias` package:

```{r install_packages}
devtools::install_github("trias-project/trias")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(tidyr)
library(tidyverse)
library(dplyr)
library(magrittr)
library(stringr)
library(readr)
# GBIF related packages
library(rgbif)
# project package
library(trias)
# INBO theme package
library(INBOtheme)
# other packages
library(lazyeval)
library(assertthat)
library(assertable)
library(egg)
```

# Get data

```{r save_data_output}
# Get data
data <- read_tsv("../data/interim/test_data_output_checklist_indicators.tsv")
```

# Build indicators

## Introduction

New functions should be written in `trias` official project package.

In order to not consider subspecies, please take care of grouping by `species` 
(see point 8 of [#21](https://github.com/trias-project/pipeline/issues/21)).

About point 7 of [#21](https://github.com/trias-project/pipeline/issues/21), 
please see this [comment](https://github.com/trias-project/pipeline/issues/
21#issuecomment-381164874).

## Number of new introductions of alien species per year

See GitHub issue [#17](https://github.com/trias-project/pipeline/issues/17).

Create function to visualize yearly introduction evolution

```{r}
# Create function
#' Create year of introduction based Trias indicator
#'
#' @param data df, as created by preprocessing
#' @param start_year_plot Start year to show the plot, as initial years are 
#' mostly irrelevant
#' @param smooth_span Parameter for the applied Loess smoother, for  more 
#' information on the appropriate value, see...
#' @param x_scale_stepsize Parameter that indicates the stepsize of the x axis
#' @param facet_column NULL | char The column to use to create additional
#' facet wrap plots underneath the main graph. When NULL, no facet graph is
#' included.
#'
#' @return ggplot
#' @export
#'
#' @importFrom assertthat assert_that
#' @importFrom assertable assert_colnames
#' @importFrom dplyr %>% filter group_by group_by_ count ungroup
#' @importFrom ggplot2 geom_point aes xlab ylab scale_x_continuous facet_wrap
#' geom_smooth
#' @importFrom INBOtheme theme_inbo
#' @importFrom egg ggarrange
#'
#' @examples
#' indicator_introduction_year(data)
#' indicator_introduction_year(data, start_year_plot = 1940, 
#'                             smooth_span = 0.6)
indicator_introduction_year <- function(df, start_year_plot = 1920,
                                       smooth_span = .85,
                                       x_major_scale_stepsize = 10,
                                       x_minor_scale_stepsize = 5,
                                       facet_column = NULL) {
    # initial input checks 
    assert_that(is.data.frame(df))
    assert_colnames(df, c("first_observed"), only_colnames = FALSE)
    
    # first filtering of the incoming data
    data <- df %>%
      filter(!is.na(.data$first_observed)) %>%
        filter(.data$first_observed > start_year_plot)
        
    data_top_graph <- data %>% 
        group_by(.data$first_observed) %>%
        count() %>%
        ungroup()
    
    # top graph with all counts
    top_graph <- ggplot(data_top_graph, aes(x = first_observed, y = n)) +
        geom_point(stat = 'identity') +
        geom_smooth(span = smooth_span) +
        xlab("Year") +
        ylab("Number of introduced alien species") +         
        scale_x_continuous(breaks = seq(start_year_plot, 
                                     max(data_top_graph$first_observed), 
                                     x_major_scale_stepsize), 
                       limits = c(start_year_plot, 
                                  max(data_top_graph$first_observed))) +
        theme_inbo()
    
    if (is.null(facet_column)) {
      return(top_graph)
    } else {
        # check for valid facet options
        valid_facet_options <- c("family", "order", "class", "phylum", 
                                 "kingdom", "pathway_level1", "locality", 
                                 "native_range")
        facet_column <- match.arg(facet_column, valid_facet_options)        
        
        data_facet_graph <- data %>% 
            group_by_("first_observed", facet_column) %>%
            count() %>%
            ungroup()
        
        facet_graph <- ggplot(data_facet_graph, aes(x = first_observed, y = n)) +
            geom_point(stat = 'identity') +
            geom_smooth(span = smooth_span) +
            facet_wrap(facet_column) +
            xlab("Year") +
            ylab("Number of introduced alien species") + 
            scale_x_continuous(breaks = seq(start_year_plot, 
                                            max(data_facet_graph$first_observed), 
                                            x_major_scale_stepsize),
                               minor_breaks = seq(start_year_plot,
                                                  max(data_facet_graph$first_observed), 
                                            x_minor_scale_stepsize),
                               limits = c(start_year_plot, 
                                          max(data_facet_graph$first_observed))) +
            theme_inbo()
         
        ggarrange(top_graph, facet_graph)
    }
 }
```

Set start_year and loess_span if you want
Specify this in the function so the function knows that it shouldn't use the 
default settings (that were specified above)
```{r, fig.width=10, fig.height=10}
# Example
start_year_plot <- 1920
loess_span <- .85
x_major_scale_stepsize <- 10
x_minor_scale_stepsize <- 2.5
indicator_introduction_year(data, start_year_plot = start_year_plot, 
                            smooth_span = loess_span,
                            x_major_scale_stepsize = x_major_scale_stepsize,
                            x_minor_scale_stepsize = x_minor_scale_stepsize,
                            facet_column = "kingdom")
```



