---
title: "Indicator: Cumulative number of alien species"
author:
- Damiano Oldoni
- Sander Devisscher
- Stijn Vanhoey
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to build indicators of Invasive Alien Species based on checklist data. In particular, this document takes into account:

1. pathways associated with alien species introductions (see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
```

Install `trias` package:

```{r install_packages}
devtools::install_github("trias-project/trias@add_functions_checklist_indicators")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(tidyr)
library(dplyr)
library(magrittr)
library(stringr)
library(readr)
library(ggplot2)
# GBIF related packages
library(rgbif)
# project package
library(trias)
#other pckages
library(lazyeval)
library(INBOtheme)
library(assertthat)
library(assertable)
library(egg)
```

# Get data

```{r save_data_output}
# Get data
data <- read_tsv("../data/interim/test_data_output_checklist_indicators.tsv")
```

# Cumulative number of introductions of alien species

We use function `indicator_total_year()` from `trias` library.

We show facetted data by kingdom:

```{r facet_column}
facet_column = "kingdom"
```

```{r fig.width=10, fig.height=15}
start_year_plot <- 1920
x_major_scale_stepsize <- 25
x_minor_scale_stepsize <- 12.5

indicator_total_year(data, facet_column = facet_column)
```

Version without facets:

```{r fig.width=10, fig.height=5}
indicator_total_year(data, start_year_plot, x_major_scale_stepsize)
```

