---
title: "Indicator: Pathways associated with alien species introductions"
author:
- Damiano Oldoni
- Add your name
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

This document describes how to build indicators of Invasive Alien Species based on checklist data. In particular, this document takes into account:

1. pathways associated with alien species introductions (see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install `trias` package:

```{r install_packages, eval=FALSE}
devtools::install_github("trias-project/trias")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(tidyr)
library(dplyr)
library(magrittr)
library(stringr)
library(readr)
library(ggplot2)
# GBIF related packages
library(rgbif)
# project package
library(trias)
#other pckages
library(lazyeval)
library(INBOtheme)
```

# Get data

```{r save_data_output}
data <- read_tsv("../data/interim/test_data_output_checklist_indicators.tsv")

```

# Build indicators

## Introduction

New functions should be written in `trias` official project package.

In order to not consider subspecies, please take care of grouping by `species` (see point 8 of [#21](https://github.com/trias-project/pipeline/issues/21)).

About point 7 of [#21](https://github.com/trias-project/pipeline/issues/21), please see this [comment](https://github.com/trias-project/pipeline/issues/21#issuecomment-381164874).

## Pathways associated with alien species introductions

see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19)

### Preprocessing

We focus on Plants and Animals, extract the pathways and count for each pathway:

```{r}
df_pathway <- data  %>%
    filter(kingdom %in% c('Plantae', 'Animalia')) %>%
    filter(type == 'pathway') %>%
    group_by(description, pathway_level1, 
            pathway_level2, kingdom) %>%
    count() %>%
    ungroup()
head(df_pathway)
```

### Figure representation

With the [existing representation](https://github.com/inbo/natuurindicatoren/blob/master/src/bedreiging-door-nieuwe-uitheemse-diersoorten.Rmd) as a reference, we can recreate the figure using the updated data:

```{r pathway-figure, fig.width=12, fig.height=12}
# uppercase for the pathway main categories
df_pathway <- mutate_at(df_pathway, .cols = c("pathway_level1"), .funs = toupper)

ggplot(df_pathway, aes(x = reorder(pathway_level2, -n), 
                       y = n, fill = kingdom)) +
    geom_bar(stat = "identity", width = .65) +
    ylab("Number of species") +
    facet_grid(. ~ pathway_level1, scales = "free_x",
               space = "free_x") +
    theme_inbo(base_size = 11) +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1,
                                   vjust = 0.2, size = 15),
        axis.title.x = element_blank(),
        axis.text.y = element_text(size = 15, angle = 90,
                                   hjust = 0.5),
        axis.title.y = element_text(size = 14),
        strip.text.x = element_text(angle = 90, size = 12),
        legend.direction = "horizontal",
        legend.position = "bottom",
        legend.text = element_text(angle = 90, size = 14),
        legend.title = element_blank(),
        legend.text.align = -.6
    )
```

Only focusing on the main category:

```{r}
ggplot(df_pathway,  aes(x = reorder(pathway_level1, n, function(x){sum(x)}),
                    y = n, fill = kingdom)) +
    geom_bar(stat = "summary", fun.y = sum,
             width = .65) + 
    xlab("Pathway") +
    ylab("Number of species") + 
    coord_flip() +
    theme(legend.position = "top",
          legend.title = element_blank()
          )
```


