---
title: "Indicator: Number of new introductions of alien species per year"
author:
- Damiano Oldoni
- Yasmine Verzelen
- Stijn Van Hoey
- Tim Adriaens
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to build indicators of Invasive Alien Species based 
on checklist data. In particular, this document takes into account:

1. number of new introductions of alien species per year in Belgium (see GitHub 
issue [#17](https://github.com/trias-project/pipeline/issues/17))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries}
library(tidyverse) # To do datascience
library(tidylog) # To provide feedback on dplyr functions
library(rlang) # To use Standard Evaluation in dplyr (!!sym())
library(here) # To find files
library(trias) # To use functions developed for TrIAS
library(INBOtheme) # To use INBO theme for graphs
```

# Get data

```{r get_data_input}
data_file <- here::here(
  "data",
  "interim",
  "data_input_checklist_indicators.tsv"
)
data <- read_tsv(data_file,
  na = "",
  guess_max = 5000
)
```

# Number of new introductions of alien species per year

We use function `indicator_introduction_year` from `trias` library.

## Select data from Belgium

We select data at national level (no regional distributions):

```{r select_data_at_Belgian_level}
data <-
  data %>%
  filter(locationId == "ISO_3166:BE")
```

## Grouping by kingdom

We group data by kingdom:

```{r facet_column_kingdom}
facet_column <- "kingdom"
```

```{r graph_grouping_kingdom, fig.width=10, fig.height=10, results="hide", echo=FALSE}
data %>%
  select(key, facet_column, first_observed) %>%
  distinct() %>%
  indicator_introduction_year(
    start_year_plot = 1920,
    smooth_span = .85,
    x_major_scale_stepsize = 10,
    x_minor_scale_stepsize = 2.5,
    facet_column = facet_column
  )
```

## Grouping by species profile

We group data by habitat:

```{r facet_column_habitat}
facet_column <- "habitat"
```

```{r graph_grouping_habitat, fig.width=10, fig.height=10, results="hide", echo=FALSE}
data %>%
  select(key, facet_column, first_observed) %>%
  distinct() %>%
  indicator_introduction_year(
    start_year_plot = 1920,
    smooth_span = .85,
    x_major_scale_stepsize = 10,
    x_minor_scale_stepsize = 2.5,
    facet_column = facet_column
  )
```

# Checklist indicators for Living Planet Index

In this section we make little changes to the indicator graphs for publication on the Living Planet Index.

## Grouping by kingdom

We group data by kingdom:

```{r facet_column_lpi_kingdom_lpi}
facet_column <- "kingdom"
```

We don't take into account Fungi and Chromista due to a lack of comprehensive data on those kingdoms:

```{r graph_grouping_kingdom_LPI, fig.width=10, fig.height=10, results="hide", echo=FALSE}
data %>%
  select(key, facet_column, first_observed) %>%
  distinct() %>%
  filter(!(!!sym(facet_column)) %in% c("Chromista", "Fungi")) %>%
  indicator_introduction_year(
    start_year_plot = 1920,
    smooth_span = .85,
    x_major_scale_stepsize = 10,
    x_minor_scale_stepsize = 2.5,
    facet_column = facet_column
  )
```

## Grouping by species profile

```{r facet_column_lpi_habitat}
facet_column <- "habitat"
```

We don't take into account combined habitats `freshwater|terrestrial` and `marine|freshwater` due to a lack of comprehensive data on those habitats:

```{r graph_grouping_habitat_LPI, fig.width=10, fig.height=10, results="hide", echo=FALSE}
data %>%
  select(key, facet_column, first_observed) %>%
  distinct() %>%
  filter(!(!!sym(facet_column)) %in% c(
    "freshwater|terrestrial",
    "marine|freshwater",
    NA_character_
  )) %>%
  indicator_introduction_year(
    start_year_plot = 1920,
    smooth_span = .85,
    x_major_scale_stepsize = 10,
    x_minor_scale_stepsize = 2.5,
    facet_column = facet_column
  )
```
