---
title: "Indicator: Number of new introductions of alien species per year"
author:
- Damiano Oldoni
- Yasmine Verzelen
- Stijn Van Hoey
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to build indicators of Invasive Alien Species based 
on checklist data. In particular, this document takes into account:

1. number of new introductions of alien species per year in Belgium (see GitHub 
issue [#17](https://github.com/trias-project/pipeline/issues/17))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install `trias` package:

```{r install_packages}
devtools::install_github("trias-project/trias")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(readr)
# Project package
library(trias)
```

# Get data

```{r save_data_output}
# Get data
data <- read_tsv("../data/interim/test_data_output_checklist_indicators.tsv")
```

# Number of new introductions of alien species per year

We use function `indicator_introduction_year()` from `trias` library.

We group data by kingdom:

```{r facet_column}
facet_column = "kingdom"
```


```{r, fig.width=10, fig.height=10, results="hide"}
indicator_introduction_year(data, start_year_plot = 1920, 
                            smooth_span = .85,
                            x_major_scale_stepsize = 10,
                            x_minor_scale_stepsize = 2.5,
                            facet_column = "kingdom")
```
