---
title: "Indicator: Cumulative number of alien species"
author:
- Damiano Oldoni
- Sander Devisscher
- Stijn Vanhoey
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to build indicators of Invasive Alien Species based on checklist data. In particular, this document takes into account:

1. pathways associated with alien species introductions (see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
```

Install `trias` package:

```{r install_packages}
devtools::install_github("trias-project/trias@add_functions_checklist_indicators")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(readr)
# project package
library(trias)
```

# Get data

```{r save_data_output}
data <- read_tsv("../data/interim/test_data_output_checklist_indicators.tsv")
```

# Cumulative number of introductions of alien species

We use function `indicator_total_year()` from `trias` library.

```{r fig.width=10, fig.height=15, results = "hide"}
indicator_total_year(data, facet_column = NULL)
```

We can facet data by kingdom:

```{r facet_column}
facet_column = "kingdom"
```

Show facetted data:

```{r fig.width=10, fig.height=5, results = "hide"}
indicator_total_year(data, 
                     start_year_plot = 1920, 
                     x_major_scale_stepsize =  25, 
                     x_minor_scale_stepsize = 12.5,
                     facet_column = facet_column)
```
