---
title: "Indicator: Cumulative number of alien species"
author:
- Damiano Oldoni
- Sander Devisscher
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

This document describes how to build indicators of Invasive Alien Species based on checklist data. In particular, this document takes into account:

1. pathways associated with alien species introductions (see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install `trias` package:

```{r install_packages}
devtools::install_github("trias-project/trias")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(tidyr)
library(dplyr)
library(magrittr)
library(stringr)
library(readr)
library(ggplot2)
# GBIF related packages
library(rgbif)
# project package
library(trias)
#other pckages
library(lazyeval)
library(INBOtheme)

```

# Get data

```{r save_data_output}
data <- read_tsv("../data/interim/test_data_output_checklist_indicators.tsv")

```

# Build indicators

## Introduction

New functions should be written in `trias` official project package.

In order to not consider subspecies, please take care of grouping by `species` (see point 8 of [#21](https://github.com/trias-project/pipeline/issues/21)).

About point 7 of [#21](https://github.com/trias-project/pipeline/issues/21), please see this [comment](https://github.com/trias-project/pipeline/issues/21#issuecomment-381164874).

## Cumulative number of alien species

see GitHub issue [#20](https://github.com/trias-project/pipeline/issues/20)

```{r Determine presence per species(key) per year}

unique_df <- data %>%
 filter(!is.na(startDate)) %>%
 distinct(key, startDate, endDate)

# Create yearslist using startDate and endDate 
counts_ias <- data.frame(year = c(min(unique_df$startDate, na.rm = TRUE):
                                 max(unique_df$endDate, na.rm = TRUE))) %>%
 as_tibble()

count_presence <- function(year, df) {
 df %>% filter(startDate <= year & endDate >= year) %>%
   nrow() %>% as_tibble() %>% rename(counts = value)
}

counts_ias %<>%
 rowwise() %>%
 do_(interp(~ count_presence(year = .$year, df = unique_df))) %>%
 bind_cols(counts_ias) %>% select(year, everything())
```

```{r Calculate time sections}

counts_ias$decade <- counts_ias$year - counts_ias$year %% 10


```



```{r Basic english Graphs}

counts_ias %>% ggplot( aes(x = year, 
                            y = counts)) +
                geom_line() +
                xlab("Year") +
                ylab("Number of alien species (cumulative sum)") +
                theme(strip.text.y = element_text(size = 16)) +
                scale_x_discrete(labels = seq(1200, 2030, 15)) + 
                scale_x_continuous(breaks = seq(1200, 2010, 50)) +
                theme_inbo(base_size = 14)

counts_ias %>% ggplot( aes(x = year, 
                            y = counts)) +
                geom_bar(stat = "identity") +
                xlab("Year") +
                ylab("Number of alien species (cumulative sum)") +
                theme(strip.text.y = element_text(size = 16)) +
                scale_x_discrete(labels = seq(1200, 2030, 15)) + 
                scale_x_continuous(breaks = seq(1200, 2010, 50)) +
                theme_inbo(base_size = 14)
```

```{r Basic dutch Graph}

counts_ias %>% ggplot( aes(x = year, 
                            y = counts)) +
                geom_line() +
                xlab("Jaar") +
                ylab("Aantal uitheemse soorten (cumulatieve som)") +
                theme(strip.text.y = element_text(size = 16)) +
                scale_x_discrete(labels = seq(1200, 2030, 15)) + 
                scale_x_continuous(breaks = seq(1200, 2010, 50))

```

