---
title: "Indicator: Cumulative number of alien species"
author:
- Damiano Oldoni
- Sander Devisscher
- Stijn Van Hoey
- Tim Adriaens
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to get the cumulative number of taxa identified as alien in Belgium.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
```

Load libraries:

```{r load_libraries}
library(tidyverse)  # To do datascience
library(tidylog)    # To provide feedback on dplyr functions
library(here)       # To find files
library(trias)      # To use functions developed for TrIAS
library(INBOtheme)  # To use INBO theme for graphs
```

# Get data

```{r read_data_in}
data_file <- here::here("data",
                        "interim",
                        "data_input_checklist_indicators.tsv")
data <- read_tsv(data_file,
                 na = "",
                 guess_max = 5000)
```

# Cumulative number of introductions of alien species

We use function `indicator_total_year` from `trias` library.

## Select taxa at Belgian level

We select data at national level (no regional distribution):

```{r select_data_at_Belgian_level}
data <-
  data %>%
  filter(locationId == "ISO_3166-2:BE")
```

## Grouping by kingdom

We group data by kingdom:

```{r facet_column}
facet_column = "kingdom"
```

```{r fig.width=10, fig.height=5, results = "hide", echo=FALSE, warning=FALSE}
data %>%
  select(speciesKey, facet_column, first_observed) %>%
  distinct() %>%
  indicator_total_year(start_year_plot = 1920,
                       x_major_scale_stepsize =  25,
                      x_minor_scale_stepsize = 12.5,
                      facet_column = facet_column)
```
