---
title: "Indicator: Pathways associated with alien species introductions"
author:
- Damiano Oldoni
- Stijn Van Hoey
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to build indicators of Invasive Alien Species based on checklist data. In particular, this document takes into account:

1. pathways associated with alien species introductions (see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
```

Install `trias` package:

```{r install_packages, eval=FALSE}
devtools::install_github("trias-project/trias")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
# library(tidyr)
library(dplyr)
library(purrr)
library(stringr)
library(readr)
# project package
library(trias)
#other pckages
library(kableExtra)
```

# Get data

```{r save_data_output}
data <- read_tsv("../data/interim/test_data_output_checklist_indicators.tsv")
```

# Build pathway indicator

## Introduction

New functions should be afterwards moved to `trias` official project package.

About point 7 of [#21](https://github.com/trias-project/pipeline/issues/21), please see this [comment](https://github.com/trias-project/pipeline/issues/21#issuecomment-381164874).

In order to not consider subspecies, please take care of grouping by `species` (see point 8 of [#21](https://github.com/trias-project/pipeline/issues/21)).

## Pathways associated with alien species introductions

See GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19).

If animal species are present, separate them in two groups: vertebrates ( `Chordata`) and invertebrates (`Not Chordata`):

```{r find_categories_in_data}
categories <- data %>% distinct(kingdom) %>%
  rename(category = kingdom)
if ("Animalia" %in% categories$category) {
  if ("Chordata" %in% data$phylum) {
    categories <- categories %>% 
      bind_rows(data.frame(category = "Chordata",
                           stringsAsFactors = FALSE))
  }
  if (data %>% 
      filter(kingdom == "Animalia") %>%
      filter(! phylum == "Chordata") %>%
      distinct(phylum) %>% nrow() > 0) {
    categories <- categories %>%
      bind_rows(data.frame(category = "Not Chordata",
                           stringsAsFactors = FALSE))
  }
  if (any(c("Chordata", "Not Chordata") %in% categories$category)) {
    categories <- categories %>%
      filter(category != "Animalia")
  }
}
```

Function  `get_table_pathways()` for creating a table based on checklist data, chosen species category and number of species returned as example per each pathway modality.

```{r pathway_table_function}
#' Pathway count indicator figure
#' 
#' @param data data.frame According to the specification of the Trias pipeline
#' @param category character One of the kingdoms as given in GBIF: \itemize{ \item{"Plantae"} \item{Animalia} \item{} \item{"Fungi"} \item{"Chromista"} \item{"Archaea"} \item{Bacteria} \item{Protozoa} \item{Viruses} \item{incertae sedis} } It can also be one of the following not kingdoms: #'\itemize{\item{Chordata} \item{Not Chordata} }
#' @param n_species integer The maximum number of species to return as examples per pathway. For groups with less species than \code{n_species}, all species are given. Default: 5.
#' 
#' @return a data.frame
#' @export
#' 
#' @examples
#' indicator_pathway(data, "Plantae", n_species = 8)
get_table_pathways <- function(data, category, n_species = 5) {
  
  # handle asymmetric cateogry system (Chordata, Not Chordta are not kingdoms) 
  if (!category %in% c("Chordata", "Not Chordata")) {
    filtered_data <- data %>% filter(kingdom == category)
  } else {
    if (category == "Chordata") {
      filtered_data <- data %>% filter(phylum == category)
    } else {
      filtered_data <- data %>% 
        filter(kingdom == "Animalia") %>%
        filter(phylum != category)
    }
  }
  
  # Create groups basd on pathway level1 and level2
  preprocess_data <- filtered_data %>% 
  distinct(scientificName, species, genus, family, order, 
         class, phylum, kingdom, speciesKey, country,
         pathway_level1, pathway_level2) %>%
  mutate(pathway_level1 = ifelse(is.na(pathway_level2), 
                                 NA, .$pathway_level1)) %>%
  group_by(pathway_level1, pathway_level2)
  
  # Assess size of sample per group
  pathway_data <- preprocess_data %>% 
    count() %>% 
    rowwise() %>%
    mutate(size_sample = ifelse(n > n_species,
                                n_species, n))
  # Make df with sample species
  samples <- pmap_dfr(list(pathway_data$pathway_level1,
                        pathway_data$pathway_level2,
                        pathway_data$size_sample),
                   function(p1, p2, s) {
                     set_species <- if (!is.na(p1)) {
                       preprocess_data %>%
                         filter(pathway_level1 == p1) %>% 
                         filter(pathway_level2 == p2)
                     } else {
                       preprocess_data %>%
                         filter(is.na(pathway_level1))
                     }
                     examples <- sample_n(set_species,
                                          s) %>% pull(species) 
                     data.frame(examples = str_c(examples, collapse = ", "),
                                stringsAsFactors = FALSE) %>% 
                       as_tibble() %>%
                       mutate(pathway_level1 = p1,
                              pathway_level2 = p2)
                   })
  
  # Join pathways and samples together
  pathway_data <- left_join(pathway_data, samples,
                            by = c("pathway_level1", "pathway_level2")) %>%
    select(-size_sample)
  
  # Create output table (untidy)
  pathway_data <- pathway_data %>%
    mutate(pathway_level1 = ifelse(!is.na(pathway_level1),
                                   pathway_level1, "unknown"),
           pathway_level2 = ifelse(pathway_level1 == "unknown",
                                   "",
                                   pathway_level2)) %>%
    mutate(examples = ifelse(!is.na(examples),
                             examples,
                             str_c(sample_n(preprocess_data %>% 
                                        filter(is.na(pathway_level1) |
                                                 is.na(pathway_level1)),
                                        n)$species,
                                   collapse = ", "))) %>%
    ungroup()
  
  pathway_data$pathway_level1[
    duplicated(pathway_data$pathway_level1) == TRUE] <- ""
  
  return(pathway_data)
}
```

Number of species returned in table as examples of the group:

```{r set_n_species}
n_species <- 5
```

Pathways tables

```{r make_tables}
tables <- map(categories$category, ~ get_table_pathways(data, ., n_species))
names(tables) <- categories$category
```

## Pathways per category

Intentionality of introduction is mapped by cell color background: white for intentional introduction, grey for unintentional introduction.

### Plantae

Pathways for Plantae:

```{r view_tables, echo = FALSE, 'asis' = TRUE}
if ("Plantae" %in% categories$category) {
  kable(tables$Plantae) %>% 
    row_spec(which(tables$Plantae$pathway_level1 == "release"), 
           background = "white") %>%
    row_spec(which(tables$Plantae$pathway_level1 != "release"), 
           background = "#faf0e6") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "10cm")
}
```

### Vertebrates

Pathways for vertebrates:

```{r view_table_Chordata, echo = FALSE}
if ("Chordata" %in% categories$category) {
    kable(tables$Chordata) %>% 
    row_spec(which(tables$Chordata$pathway_level1 == "release"), 
           background = "white") %>%
    row_spec(which(tables$Chordata$pathway_level1 != "release"), 
           background = "#faf0e6") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "10cm")
}
```

### Invertebrates

Pathways for invertebrates:

```{r view_table_NotChordata, echo = FALSE}
if ("Not Chordata" %in% categories$category) {
  kable(tables$`Not Chordata`) %>% 
    row_spec(which(tables$`Not Chordata`$pathway_level1 == "release"), 
           background = "white") %>%
    row_spec(which(tables$`Not Chordata`$pathway_level1 != "release"), 
           background = "#faf0e6") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "10cm")
}
```

### Fungi

Pathways for Fungi:

```{r view_table_Fungi, echo = FALSE}
if ("Fungi" %in% categories$category) {
  tables$Fungi
}
```

### Bacteria

Pathways for Bacteria:

```{r view_table_Bacteria, echo = FALSE}
if ("Bacteria" %in% categories$category) {
  tables$Bacteria
}
```

Add sections for other kingdoms if needed.
