---
title: "Indicator: Pathways associated with alien species introductions"
author:
- Damiano Oldoni
- Stijn Van Hoey
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

This document describes how to build indicators of Invasive Alien Species based on checklist data. In particular, this document takes into account:

1. pathways associated with alien species introductions (see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install `trias` package:

```{r install_packages, eval=FALSE}
devtools::install_github("trias-project/trias")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(tidyr)
library(dplyr)
library(magrittr)
library(stringr)
library(readr)
library(ggplot2)
# GBIF related packages
library(rgbif)
# project package
library(trias)
#other pckages
library(lazyeval)
library(INBOtheme)
library(stringr)
```

# Get data

```{r save_data_output}
data <- read_tsv("../data/interim/test_data_output_checklist_indicators.tsv")

```

# Build indicators

## Introduction

New functions should be written in `trias` official project package.

In order to not consider subspecies, please take care of grouping by `species` (see point 8 of [#21](https://github.com/trias-project/pipeline/issues/21)).

About point 7 of [#21](https://github.com/trias-project/pipeline/issues/21), please see this [comment](https://github.com/trias-project/pipeline/issues/21#issuecomment-381164874).

## Pathways associated with alien species introductions

See GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19). Ths functionality is scripted inside a function, with the two usecase taken into account:

1. using the subcategory and category level, making a double grouping in the graph
2. using the category level only

As the function is ad-hoc to the use-case, most options are hard-coded.

```{r}

#' Pathway count indicator figure 
#'
#' Both only the main category or using both category levels are included in the
#' graph, using the \code{subcategory} option
#'
#' @param data data.frame According to the specification of the Trias pipeline
#' @param subcategory boolean If TRUE, the subcategory levels are included in 
#' the visualisation.
#'
#' @return
#' @export
#'
#' @examples
#' indicator_pathway(data)
#' indicator_pathway(data, subcategory = TRUE)
indicator_pathway <- function(data, subcategory = TRUE) {
    
    # focus on pathway data counts
    df_pathway <- data  %>%
        filter(kingdom %in% c('Plantae', 'Animalia')) %>%
        group_by(pathway_level1, 
                pathway_level2, kingdom) %>%
        count() %>%
        ungroup()
    
    # uppercase on category names for figure styling
    df_pathway <- mutate_at(df_pathway, 
                            .cols = c("pathway_level1"), 
                            .funs = toupper)
    # replace underscore by spaces
    df_pathway <- df_pathway %>%
        mutate(pathway_level2 = str_replace(pathway_level2, "_", " "))
        
    if (subcategory) {
        ggplot(df_pathway, aes(x = reorder(pathway_level2, -n), 
                               y = n, fill = kingdom)) +
            geom_bar(stat = "identity", width = .65) +
            ylab("Number of species") +
            facet_grid(. ~ pathway_level1, scales = "free_x",
                       space = "free_x") +
            theme_inbo(base_size = 14) +
            theme(
                axis.text.x = element_text(angle = 90, hjust = 1,
                                           vjust = 0.2, size = 15),
                axis.title.x = element_blank(),
                axis.text.y = element_text(size = 15, angle = 90,
                                           hjust = 0.5),
                axis.title.y = element_text(size = 14),
                strip.text.x = element_text(angle = 90, size = 12),
                legend.direction = "horizontal",
                legend.position = "right",
                legend.text = element_text(angle = 90, size = 14),
                legend.title = element_blank(),
                legend.text.align = -.6)
    } else {
        ggplot(df_pathway,  aes(x = reorder(pathway_level1, n, function(x){sum(x)}),
                            y = n, fill = kingdom)) +
            geom_bar(stat = "summary", fun.y = sum,
                     width = .65) + 
            xlab("Pathway") +
            ylab("Number of species") + 
            coord_flip() +
            theme_inbo(base_size = 14) +
            theme(legend.position = "top",
                  legend.title = element_blank(),
                  legend.text = element_text(size = 14))
    }
}
```

Example usage of the function when using both category levels:

```{r pathway-figure, fig.width=12, fig.height=10}
indicator_pathway(data, subcategory = TRUE)
```

Or when only interested in the main category in the graph:

```{r pathway-figure, fig.width=8, fig.height=6}
indicator_pathway(data, subcategory = FALSE)
```





