---
title: "Indicator: Cumulative number of alien species"
author:
- Damiano Oldoni
- Sander Devisscher
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

This document describes how to build indicators of Invasive Alien Species based on checklist data. In particular, this document takes into account:

1. pathways associated with alien species introductions (see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install `trias` package:

```{r install_packages}
devtools::install_github("trias-project/trias")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(tidyr)
library(dplyr)
library(magrittr)
library(stringr)
library(readr)
library(ggplot2)
# GBIF related packages
library(rgbif)
# project package
library(trias)
#other pckages
library(lazyeval)
library(INBOtheme)

```

# Get data

```{r save_data_output}
data <- read_tsv("../data/interim/test_data_output_checklist_indicators.tsv")

```

# Build indicators

## Introduction

New functions should be written in `trias` official project package.

In order to not consider subspecies, please take care of grouping by `species` (see point 8 of [#21](https://github.com/trias-project/pipeline/issues/21)).

About point 7 of [#21](https://github.com/trias-project/pipeline/issues/21), please see this [comment](https://github.com/trias-project/pipeline/issues/21#issuecomment-381164874).

## Cumulative number of alien species

see GitHub issue [#20](https://github.com/trias-project/pipeline/issues/20)

```{r determine_presence_per_specieskey_per_year}

unique_df <- data %>%
                  filter(!is.na(startDate)) %>%
                  distinct(key, startDate, endDate)

# Create yearslist using startDate and endDate 
counts_ias <- data.frame(year = c(min(unique_df$startDate, na.rm = TRUE):
                                  max(unique_df$endDate, na.rm = TRUE))) %>%
                                  as_tibble()

count_presence <- function(year, df) {
                  df %>% filter(startDate <= year & endDate >= year) %>%
                  nrow() %>% as_tibble() %>% rename(counts = value)
                  }

counts_ias %<>%
           rowwise() %>%
           do_(interp(~ count_presence(year = .$year, df = unique_df))) %>%
           bind_cols(counts_ias) %>% select(year, everything())
```

```{r calculate_time_sections}
#calculate the decade 
counts_ias$decade <- counts_ias$year - counts_ias$year %% 10
```

```{r calculate_min_max_date}
#calculate the minimum and maximum date
#mindate is calculated to the decade
minDate <- min(counts_ias$year) - min(counts_ias$year) %% 10
maxDate <- max(counts_ias$year)
```


```{r basic_english_line_graphs, fig.width=8, fig.height=8}

counts_ias %>% ggplot(aes(x = year, 
                          y = counts)) +
                geom_line() +
                xlab("Year") +
                ylab("Number of alien species (cumulative sum)") +
                scale_x_discrete(labels = seq(minDate, maxDate, 15)) + 
                scale_x_continuous(breaks = seq(minDate, maxDate, 50)) +
                theme(strip.text.y = element_text(size = 16)) +
                theme_inbo(base_size = 14)

```

```{r basic_english_bar_graphs, fig.width=8, fig.height=8}
counts_ias %>% ggplot(aes(x = year, 
                          y = counts)) +
                geom_bar(stat = "identity") +
                xlab("Year") +
                ylab("Number of alien species (cumulative sum)") +
                scale_x_discrete(labels = seq(minDate, maxDate, 15)) + 
                scale_x_continuous(breaks = seq(minDate, maxDate, 50)) +
                theme(strip.text.y = element_text(size = 16)) +
                theme_inbo(base_size = 14)
```

```{r basic_dutch_line_graph, fig.width=8, fig.height=8}
counts_ias %>% ggplot( aes(x = year, 
                            y = counts)) +
                geom_line() +
                xlab("Jaar") +
                ylab("Aantal uitheemse soorten (cumulatieve som)") +
                scale_x_discrete(labels = seq(minDate, maxDate, 15)) + 
                scale_x_continuous(breaks = seq(minDate, maxDate, 50)) +
                theme(strip.text.y = element_text(size = 16)) +
                theme_inbo(base_size = 14)
```

```{r basic_dutch_bar_graph, fig.width=8, fig.height=8}
counts_ias %>% ggplot(aes(x = year, 
                          y = counts)) +
                geom_line() +
                xlab("Jaar") +
                ylab("Aantal uitheemse soorten (cumulatieve som)") +
                scale_x_discrete(labels = seq(minDate, maxDate, 15)) + 
                scale_x_continuous(breaks = seq(minDate, maxDate, 50)) +
                theme(strip.text.y = element_text(size = 16)) +
                theme_inbo(base_size = 14)
```

