---
title: "Indicator: Pathways associated with alien species introductions"
author:
- Damiano Oldoni
- Stijn Van Hoey
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to build indicators of Invasive Alien Species based on checklist data. In particular, this document takes into account:

1. pathways associated with alien species introductions (see GitHub isue [#19](https://github.com/trias-project/pipeline/issues/19))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries}
library(dplyr)
library(purrr)
library(stringr)
library(readr)
library(trias)
library(kableExtra)
library(here)
```

# Get data

```{r read_data_for_building_indicators}
data_file <- here("data","output", "data_input_checklist_indicators.tsv")
data <- read_tsv(data_file, na = "")
```

# Build pathway indicator

## Find categories

We divide the species in categories. Each kingdom, except _Animalia_, is a category. If animal species are present, they are separated in two categories: vertebrates (phylum `Chordata`) and invertebrates (phylum other than `Chordata`, defined as `Not Chordata`).

Categories present in our checklist data:

```{r find_categories_in_data, echo=FALSE}
categories <- data %>% distinct(kingdom) %>%
  rename(category = kingdom)
if ("Animalia" %in% categories$category) {
  if ("Chordata" %in% data$phylum) {
    categories <- categories %>% 
      bind_rows(data.frame(category = "Chordata",
                           stringsAsFactors = FALSE))
  }
  if (data %>% 
      filter(kingdom == "Animalia") %>%
      filter(!phylum == "Chordata") %>%
      distinct(phylum) %>% nrow() > 0) {
    categories <- categories %>%
      bind_rows(data.frame(category = "Not Chordata",
                           stringsAsFactors = FALSE))
  }
  if (any(c("Chordata", "Not Chordata") %in% categories$category)) {
    categories <- categories %>%
      filter(category != "Animalia")
  }
}
categories
```

## Pathways associated with alien species introductions

Function `get_table_pathways()` from `trias` library is used for creating tables based on checklist data. Species category and number of species returned as example per each pathway modality are provided. 

Number of species returned in table as examples of the group:

```{r set_n_species}
n_species <- 5
```

We limit to pathways of taxa introduced from 2000.

```{r year_of_introduction}
start_year <- 2000
```

Create tables:

```{r make_tables_all_data, results="hide"}
tables <- map(categories$category, ~ 
                get_table_pathways(data, 
                                   category = ., 
                                   from = start_year,
                                   n_species = n_species))
names(tables) <- categories$category
```

## Pathways per category

Intentionality of introduction is mapped by cell color background: white for intentional introduction (`release`), light orange for unintentional introduction.

### Plantae

Pathways associated to plants (category `Plantae`):

```{r view_tables, echo = FALSE}
if ("Plantae" %in% categories$category) {
  kable(tables$Plantae) %>% 
    row_spec(which(tables$Plantae$pathway_level1 == "release"), 
           background = "white") %>%
    row_spec(which(tables$Plantae$pathway_level1 != "release"), 
           background = "#faf0e6") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "10cm")
}
```

### Vertebrates

Pathways associated to vertebrates (category `Chordata`):

```{r view_table_Chordata, echo = FALSE}
if ("Chordata" %in% categories$category) {
    kable(tables$Chordata) %>% 
    row_spec(which(tables$Chordata$pathway_level1 == "release"), 
           background = "white") %>%
    row_spec(which(tables$Chordata$pathway_level1 != "release"), 
           background = "#faf0e6") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "10cm")
}
```

### Invertebrates

Pathways associated to invertebrates (category `Not Chordata`):

```{r view_table_NotChordata, echo = FALSE}
if ("Not Chordata" %in% categories$category) {
  kable(tables$`Not Chordata`) %>% 
    row_spec(which(tables$`Not Chordata`$pathway_level1 == "release"), 
           background = "white") %>%
    row_spec(which(tables$`Not Chordata`$pathway_level1 != "release"), 
           background = "#faf0e6") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "10cm")
}
```

### Fungi

Pathways associated to fungi (category `Fungi`):

```{r view_table_Fungi, echo = FALSE}
if ("Fungi" %in% categories$category) {
  kable(tables$Fungi) %>% 
    row_spec(which(tables$Fungi$pathway_level1 == "release"), 
           background = "white") %>%
    row_spec(which(tables$Fungi$pathway_level1 != "release"), 
           background = "#faf0e6") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "10cm")
}
```

Note: add sections for other kingdoms if needed in the future.

## Pathways related to policy relevant species

It is important to focus our attention also to a intercategory group of species which is of particular policy importance: the *List of Invasive Alien Species of Union concern* (see more info about [here](http://ec.europa.eu/environment/nature/invasivealien/list/index_en.htm)).

We import this list of species:

```{r}
eu_concern_file <- here("data","input", "eu_concern_species.tsv")
eu_concern_species <- read_tsv(eu_concern_file)
```

Check which species from this subgroup are not present in `data`:

```{r check_presence_eu_species}
eu_concern_species %>% filter(!backbone_taxonKey %in% data$nubKey)
```

Categories present in this subgroup:

```{r categories_eu_concern, echo = FALSE}
categories_eu_concern <- data %>%
  filter(nubKey %in% eu_concern_species$backbone_taxonKey) %>%
  distinct(kingdom) %>%
  rename(category = kingdom)
if ("Animalia" %in% categories_eu_concern$category) {
  if ("Chordata" %in% data$phylum) {
    categories_eu_concern <- categories_eu_concern %>% 
      bind_rows(data.frame(category = "Chordata",
                           stringsAsFactors = FALSE))
  }
  if (data %>% 
      filter(kingdom == "Animalia") %>%
      filter(!phylum == "Chordata") %>%
      distinct(phylum) %>% nrow() > 0) {
    categories_eu_concern <- categories_eu_concern %>%
      bind_rows(data.frame(category = "Not Chordata",
                           stringsAsFactors = FALSE))
  }
  if (any(c("Chordata", "Not Chordata") %in% categories_eu_concern$category)) {
    categories_eu_concern <- categories_eu_concern %>%
      filter(category != "Animalia")
  }
}
categories_eu_concern
```

As done before, we limit to pathways of taxa introduced from 2000.

Create tables:

```{r make_tables_eu_concern, results = "hide"}
tables_eu_concern <- map(categories_eu_concern$category, ~ 
  get_table_pathways(data %>%
                       filter(nubKey %in% eu_concern_species$backbone_taxonKey), 
                     category = ., 
                     from = start_year,
                     n_species = n_species))
names(tables_eu_concern) <- categories_eu_concern$category
```

### Plantae

Pathways associated to plants (category `Plantae`):

```{r eu_concern_plantae, echo = FALSE}
if ("Plantae" %in% categories_eu_concern$category) {
  kable(tables_eu_concern$Plantae) %>% 
    row_spec(which(tables_eu_concern$Plantae$pathway_level1 == "release"), 
           background = "white") %>%
    row_spec(which(tables_eu_concern$Plantae$pathway_level1 != "release"), 
           background = "#faf0e6") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "10cm")
}
```

### Chordata

Pathways associated to vertebrates (category `Chordata`):

```{r eu_concern_chordata, echo = FALSE}
if ("Chordata" %in% categories_eu_concern$category) {
  kable(tables_eu_concern$Chordata) %>% 
    row_spec(which(tables_eu_concern$Plantae$pathway_level1 == "release"), 
           background = "white") %>%
    row_spec(which(tables_eu_concern$Chordata$pathway_level1 != "release"), 
           background = "#faf0e6") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "10cm")
}
```

### Not Chordata

Pathways associated to invertebrates (category `Not Chordata`):

```{r eu_concern_not_chordata, echo = FALSE}
if ("Not Chordata" %in% categories_eu_concern$category) {
  kable(tables_eu_concern$`Not Chordata`) %>% 
    row_spec(which(tables_eu_concern$`Not Chordata`$pathway_level1 == "release"), 
           background = "white") %>%
    row_spec(which(tables_eu_concern$`Not Chordata`$pathway_level1 != "release"), 
           background = "#faf0e6") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "10cm")
}
```
