---
title: "Indicator: Number of new introductions of alien species per year"
author:
- Damiano Oldoni
- Yasmine Verzelen
- Stijn Van Hoey
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to build indicators of Invasive Alien Species based 
on checklist data. In particular, this document takes into account:

1. number of new introductions of alien species per year in Belgium (see GitHub 
issue [#17](https://github.com/trias-project/pipeline/issues/17))

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries}
library(tidyverse)  # To do datascience
library(here)       # To find files
library(trias)      # To use functions developed for TrIAS
library(INBOtheme)  # To use INBO theme for graphs
```

# Get data

```{r get_data_input}
data <- read_tsv(here("data","output", "data_input_checklist_indicators.tsv"),
                 na = "")
```

# Number of new introductions of alien species per year

We use function `indicator_introduction_year()` from `trias` library.

## Grouping by kingdom

We group data by kingdom:

```{r facet_column}
facet_column <- "kingdom"
```

```{r, fig.width=10, fig.height=10, results="hide"}
indicator_introduction_year(data, start_year_plot = 1920,
                            smooth_span = .85,
                            x_major_scale_stepsize = 10,
                            x_minor_scale_stepsize = 2.5,
                            facet_column = facet_column)
```

## Grouping by species profile

We group data by habitat:

```{r facet_column}
facet_column <- "habitat"
```

```{r, fig.width=10, fig.height=10, results="hide"}
indicator_introduction_year(data, start_year_plot = 1920, 
                            smooth_span = .85,
                            x_major_scale_stepsize = 10,
                            x_minor_scale_stepsize = 2.5,
                            facet_column = facet_column)
```

# Checklist indicators for Living Planet Index

In this section we make little changes to the indicator graphs for publication on the Living Planet Index.

## Grouping by kingdom

We group data by kingdom:

```{r facet_column_lpi_kingdom_lpi}
facet_column <- "kingdom"
```

We don't take into account Fungi and Chromista due to a lack of comprehensive data on those kingdoms:

```{r, fig.width=10, fig.height=10, results="hide"}
indicator_introduction_year(
  data %>%
  filter(!kingdom %in% c("Chromista", "Fungi")), 
  start_year_plot = 1920, 
  smooth_span = .85,
  x_major_scale_stepsize = 10,
  x_minor_scale_stepsize = 2.5,
  facet_column = facet_column)
```

## Grouping by species profile

```{r facet_column_lpi_habitat}
facet_column <- "habitat"
```

We don't take into account combined habitats `freshwater|terrestrial` and `marine|freshwater` due to a lack of comprehensive data on those habitats:

```{r, fig.width=10, fig.height=10, results="hide"}
indicator_introduction_year(
  data %>%
  filter(!habitat %in% c("freshwater|terrestrial",
                         "marine|freshwater",
                         NA_character_)), 
  start_year_plot = 1950, 
  smooth_span = .85,
  x_major_scale_stepsize = 10,
  x_minor_scale_stepsize = 2.5,
  facet_column = facet_column)
```
