---
title: 'Emerging status: modelling'
author:
- Damiano Oldoni
- Toon Van Daele
- Tim Adriaens
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
---

This document describes the modelling to assess the emerging status of alien species.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

Load libraries:

```{r load_libraries, message = FALSE}
library(tidyverse) # To do data science
library(tidyselect) # To help tidyverse functions
library(tidylog) # To provide feedback on dplyr functions
library(rgbif) # To get information from GBIF
library(lubridate) # To work with dates
library(here) # To find files
library(trias)
library(INBOtheme) # To load graphic INBO theme
```

# Get data

We read the time series data, output of preprocessing pipeline:

```{r read_preprocessed_data, message=FALSE}
df_ts <- read_tsv(
  here::here("data", "interim", "df_timeseries.tsv"),
  na = ""
)
```

Columns `pa_obs` and `pa_cobs` indicate the presence (1) or absence (0) of the specific taxon and any other taxa within same class respectively.

Preview:

```{r preview_ts}
# Get a taxon
taxon <-
  df_ts$taxonKey[5]

# Preview
df_ts %>%
  filter(
    taxonKey == taxon,
    year %in% c(2016, 2017),
    eea_cell_code %in% c("1kmE3814N3108", "1kmE3894N3090")
  )
```

Retrieve scientific names which will be useful to better discuss the results.

```{r taxon_names, message=FALSE}
spec_names <- read_tsv(
  file = here::here("data", "interim", "timeseries_taxonomic_info.tsv"),
  na = ""
) %>%
  select(taxonKey, canonicalName) %>%
  filter(taxonKey %in% df_ts$taxonKey)
```

Lump geographic information:

```{r df_ts_compact}
df_ts_compact <-
  df_ts %>%
  group_by(taxonKey, year, classKey) %>%
  summarise(
    obs = sum(obs),
    cobs = sum(cobs),
    ncells = sum(pa_cobs),
    c_ncells = sum(pa_cobs)
  ) %>%
  ungroup()
```

We create a similar data.frame specific to protected areas (Natura2000):

```{r df_ts_compact_protected_areas}
df_ts_compact_prot_areas <-
  df_ts %>%
  filter(natura2000 == TRUE) %>%
  group_by(taxonKey, year) %>%
  summarise(
    obs = sum(obs),
    cobs = sum(cobs),
    ncells = sum(pa_cobs),
    c_ncells = sum(pa_cobs)
  ) %>%
  ungroup()
```

Add canonical names:

```{r add_canonicalName}
df_ts_compact <-
  df_ts_compact %>%
  left_join(spec_names, by = "taxonKey")
```

# Modelling

In this section we evaluate the emergenging status by applying a decision rule strategy or, where possible, a statical model called GAM (Generalized Additive Models).  For each evaluation year (see below) the output of both models is one of the following emerging status codes:

- `3`: emerging
- `2`: potentially emerging
- `1`: unclear
- `0`: not emerging

## Define evaluation period 

We define the time window (in years) we want to assess the emerging status:

```{r eval_years}
# Last evaluation year
last_year <- lubridate::year(Sys.Date()) - 2

# First evalution year
first_year <- last_year - 2

# Evaluation years
evaluation_years <- seq(first_year, last_year)
evaluation_years
```

## Decision rules

We define and apply some decision rules for occupancy and observations in all Belgium and in protected areas using function `apply_decision_rules` from project package `trias`.

### Observations

#### Belgium

We apply function `apply_decision_rules()` to observations for all Belgium:

```{r apply_decision_rules_occs_be, message=FALSE}
em_decision_rules_occs_BE <- map_dfr(
  evaluation_years,
  ~ apply_decision_rules(
    df = df_ts_compact,
    y_var = "obs",
    eval_year = .
  )
)
```

Examples:

```{r example_dec_rules_occs_BE}
em_decision_rules_occs_BE %>% head()
```

#### Protected areas

We apply function `apply_decision_rules()` to observations for protected areas:

```{r apply_decision_rules_occs_pa, message=FALSE}
em_decision_rules_occs_pa <- map_dfr(
  evaluation_years,
  ~ apply_decision_rules(
    df = df_ts_compact_prot_areas,
    y_var = "obs",
    eval_year = .
  )
)
```

Examples:

```{r example_dec_rules_occs_pa}
em_decision_rules_occs_pa %>% head()
```

### Occupancy

#### Belgium

We apply function `apply_decision_rules()` to occupancy for all Belgium:

```{r apply_decision_rules_occupancy_BE, message=FALSE}
em_decision_rules_occupancy_BE <- map_dfr(
  evaluation_years,
  ~ apply_decision_rules(
    df = df_ts_compact,
    y_var = "ncells",
    eval_year = .
  )
)
```

Examples:

```{r example_decision_rules}
em_decision_rules_occupancy_BE %>% head()
```

#### Protected areas

We apply function `apply_decision_rules()` to occupancy in protected areas:

```{r apply_dec_rules_occupancy_pa, message=FALSE}
em_decision_rules_occupancy_pa <- map_dfr(
  evaluation_years,
  ~ apply_decision_rules(
    df = df_ts_compact_prot_areas,
    y_var = "ncells",
    eval_year = .
  )
)
```

Examples:

```{r example_dr_occupancy_pa}
em_decision_rules_occupancy_pa %>% head()
```

## Generalized additive model (GAM)

We apply GAM to observations and occupancy in all Belgium and in protected areas using function `apply_gam` from project package  `trias`.

### observations

#### Belgium

Apply GAM for observation data in all Belgium:

```{r apply_gam_to_all_taxa, message=FALSE, cache=TRUE}
taxon_keys <- spec_names$taxonKey
taxon_names <- spec_names$canonicalName
gam_occs_BE <- map2(
  taxon_keys, taxon_names,
  function(t, n) {
    df_key <- df_ts_compact %>%
      filter(taxonKey == t) %>%
      filter(year <= max(evaluation_years))
    if (!is.na(classKey)) {
      results_gam <- apply_gam(
        df = df_key,
        year = "year",
        y_var = "obs",
        taxonKey = "taxonKey",
        eval_years = evaluation_years,
        type_indicator = "observations",
        taxon_key = t,
        name = n,
        baseline_var = "cobs",
        dir_name = "./data/output/GAM_outputs/",
        saveplot = TRUE,
        verbose = TRUE
      )
    } else {
      results_gam <- apply_gam(
        df = df_key,
        year = "year",
        y_var = "obs",
        taxonKey = "taxonKey",
        eval_years = evaluation_years,
        type_indicator = "observations",
        taxon_key = t,
        name = n,
        verbose = TRUE,
        saveplot = TRUE
      )
    }

    return(results_gam)
  }
)
names(gam_occs_BE) <- taxon_keys
```

Show some results:

```{r show_results_preview_occs_BE, message=FALSE}
map(
  gam_occs_BE[1:20],
  function(t) {
    if (!is.null(t$plot)) t$plot
  }
)
```

#### Protected aras

Apply GAM for observation data in protected areas:

```{r apply_gam_observations_to_all_taxa_protected_areas, message=FALSE, cache=TRUE}
taxon_keys <- spec_names$taxonKey
taxon_names <- spec_names$canonicalName
gam_occs_pa <- map2(
  taxon_keys, taxon_names,
  function(t, n) {
    df_key <- df_ts_compact_prot_areas %>%
      dplyr::filter(taxonKey == t)
    apply_gam(
      df = df_key,
      y_var = "obs",
      year = "year",
      eval_years = evaluation_years,
      type_indicator = "observations",
      taxon_key = t,
      name = n,
      baseline_var = "cobs",
      df_title = "Natura2000",
      verbose = FALSE,
      saveplot = TRUE
    )
  }
)
names(gam_occs_pa) <- taxon_keys
```

Show some results:

```{r preview_gam_occs_PA, message=FALSE}
map(
  gam_occs_pa[1:10],
  function(t) {
    if (!is.null(t$plot)) t$plot
  }
)
```

### Occupancy

#### Belgium

Apply GAM for occupancy data (number of occupied cells) in all Belgium:

```{r apply_gam_occupancy_to_all_taxa, message=FALSE, cache=TRUE}
taxon_keys <- spec_names$taxonKey
taxon_names <- spec_names$canonicalName
gam_occupancy_BE <- map2(
  taxon_keys, taxon_names,
  function(t, n) {
    df_key <- df_ts_compact %>%
      filter(taxonKey == t) %>%
      filter(year <= max(evaluation_years))
    apply_gam(
      df = df_key,
      year = "year",
      y_var = "ncells",
      eval_years = evaluation_years,
      type_indicator = "occupancy",
      taxon_key = t,
      name = n,
      baseline_var = "class_ncells",
      verbose = FALSE,
      saveplot = TRUE
    )
  }
)
names(gam_occupancy_BE) <- taxon_keys
```

Show some results:

```{r preview_gam_occupancy_BE, message=FALSE}
map(
  gam_occupancy_BE[1:10],
  function(t) {
    if (!is.null(t$plot)) t$plot
  }
)
```

#### Protected aras

Apply GAM for occupancy in protected areas:

```{r apply_gam_occupancy_to_all_taxa_protected_areas, message=FALSE, cache=TRUE}
taxon_keys <- spec_names$taxonKey
taxon_names <- spec_names$canonicalName
gam_occupancy_pa <- map2(
  taxon_keys, taxon_names,
  function(t, n) {
    df_key <- df_ts_compact_prot_areas %>%
      dplyr::filter(taxonKey == t)
    apply_gam(
      df = df_key,
      y_var = "ncells",
      year = "year",
      eval_years = evaluation_years,
      type_indicator = "occupancy",
      taxon_key = t,
      name = n,
      baseline_var = "class_ncells",
      df_title = "Natura2000",
      verbose = FALSE,
      saveplot = TRUE
    )
  }
)
names(gam_occupancy_pa) <- taxon_keys
```

Show some results:

```{r preview_gam_occupancy_PA, message=FALSE}
map(
  gam_occupancy_pa[1:10],
  function(t) {
    if (!is.null(t$plot)) t$plot
  }
)
```

# Save results

## Decision rules

Save emerging status based on decision rules:

```{r save_decisionrules_results}
write_tsv(em_decision_rules_occs_BE,
  path = here::here(
    "data", "output",
    "decision_rules_outputs",
    "output_decision_rules_occs_belgium.tsv"
  ),
  na = ""
)
write_tsv(em_decision_rules_occs_pa,
  path = here::here(
    "data", "output",
    "decision_rules_outputs",
    "output_decision_rules_occs_pa.tsv"
  ),
  na = ""
)
write_tsv(em_decision_rules_occupancy_BE,
  path = here::here(
    "data", "output",
    "decision_rules_outputs",
    "output_decision_rules_occupancy_belgium.tsv"
  ),
  na = ""
)
write_tsv(em_decision_rules_occupancy_pa,
  path = here::here(
    "data", "output",
    "decision_rules_outputs",
    "output_decision_rules_occupancy_pa.tsv"
  ),
  na = ""
)
```


## GAM models

Save complete outputs:

```{r  save_outputs}
method_em <- gam_occs_BE[[1]]$em_summary$method[1]
write_tsv(map_dfr(gam_occs_BE, function(x) {
  x$output
}),
na = "",
path = here(
  "data",
  "output",
  "GAM_outputs",
  paste0(
    "outptut_GAM_observations_belgium_",
    method_em, ".tsv"
  )
)
)

method_em <- gam_occs_pa[[1]]$em_summary$method[1]
write_tsv(map_dfr(gam_occs_pa, function(x) {
  x$output
}),
na = "",
path = here(
  "data",
  "output",
  "GAM_outputs",
  paste0(
    "outptut_GAM_observations_pa_",
    method_em, ".tsv"
  )
)
)

method_em <- gam_occupancy_BE[[1]]$em_summary$method[1]
write_tsv(map_dfr(gam_occupancy_BE, function(x) {
  x$output
}),
na = "",
path = here(
  "data",
  "output",
  "GAM_outputs",
  paste0(
    "outptut_GAM_occupancy_belgium_",
    method_em, ".tsv"
  )
)
)

method_em <- gam_occupancy_pa[[1]]$em_summary$method[1]
write_tsv(map_dfr(gam_occupancy_pa, function(x) {
  x$output
}),
na = "",
path = here(
  "data",
  "output",
  "GAM_outputs",
  paste0(
    "outptut_GAM_occupancy_pa_",
    method_em, ".tsv"
  )
)
)
```

Save summary of results limited to the years included in evaluation:

```{r save_em_summary}
method_em <- gam_occs_BE[[1]]$em_summary$method[1]
write_tsv(map_dfr(gam_occs_BE, function(x) {
  x$em_summary
}),
na = "",
path = here(
  "data",
  "output",
  "GAM_outputs",
  paste0(
    "summary_GAM_observations_belgium_",
    method_em, ".tsv"
  )
)
)

method_em <- gam_occs_pa[[1]]$em_summary$method[1]
write_tsv(map_dfr(gam_occs_pa, function(x) {
  x$em_summary
}),
na = "",
path = here(
  "data",
  "output",
  "GAM_outputs",
  paste0(
    "summary_GAM_observations_pa_",
    method_em, ".tsv"
  )
)
)

method_em <- gam_occupancy_BE[[1]]$em_summary$method[1]
write_tsv(map_dfr(gam_occupancy_BE, function(x) {
  x$em_summary
}),
na = "",
path = here(
  "data",
  "output",
  "GAM_outputs",
  paste0(
    "summary_GAM_occupancy_belgium_",
    method_em, ".tsv"
  )
)
)

method_em <- gam_occupancy_pa[[1]]$em_summary$method[1]
write_tsv(map_dfr(gam_occupancy_pa, function(x) {
  x$em_summary
}),
na = "",
path = here(
  "data",
  "output",
  "GAM_outputs",
  paste0(
    "summary_GAM_occupancy_pa_",
    method_em, ".tsv"
  )
)
)
```
