---
title: "Occurrence indicators: downscaling"
author:
- Damiano Oldoni
- Add your name
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

This document describes how to build indicators of Invasive Alien Species based on occurrence data.
The following indicators will be built:

1.
2.



# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install `trias` package:

```{r install_packages}
devtools::install_github("trias-project/trias")
```


Load libraries:

```{r load_libraries}
# Tidyverse packages
library(tidyr)
library(dplyr)
library(magrittr)
library(stringr)
library(readr)
# GBIF related packages
library(rgbif)
# project package
library(trias)
#other pckages
library(lazyeval)
```

# Load data

We can import data based on a previous download (see `./get_occurrences.Rmd` pipeline for getting occurrences data from GBIF). The download keys and DOIs are saved in file `../data/output/gbif_downloads.tsv`. A reference to input taxa is present as well.
So, we can load data related to taxa in `../data/input/species_test_invasion_curve.tsv`, by using the `gbif_download_key`, `0020843-180131172636756` (first column of `gbif_downloads.tsv`) in `rgbif` function `occ_download_get()`:

```{r load_occ_data}
gbif_download_key <- "0020843-180131172636756"
occ <- occ_download_get(key = gbif_download_key, 
                        overwrite = T, 
                        path = "../data/interim/")
fn <- "occurrence.txt"
unzip(zipfile = occ, files = fn, 
      exdir = "../data/interim")
file.rename(from = "../data/interim/occurrence.txt", 
            to = "../data/interim/occurrence_species_test_invasion_curve.txt")
occ_df <- read_delim(
  file = "../data/interim/occurrence_species_test_invasion_curve.txt", "\t",
  escape_double = FALSE, trim_ws = TRUE)
```

# Downscaling

Put your workflow here.
